---
title: "Athlete Activities Cleaning"
author: "Peter Bonner"
date: "15/04/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, warning = FALSE,
                      message = FALSE,echo = TRUE, dpi = 180,
                      fig.width = 8, fig.height = 5)


```

## Libraries
```{r}
library(tidyverse)
library(janitor)
library(readxl)
library(lubridate)
library(ggthemr)
ggthemr("fresh")


```

## Data Cleaning


##### Read data
```{r read-data}
activities <- read_csv("data/activities.csv") %>%
  filter(sport == "Bike")

```



* add in training season and training phases:

```{r}

# Create date variables

activities_var <- activities %>%
  mutate(date = as_date(date),
         year = year(date),
         # week = week(date),
         month = as.numeric(format(date, "%m"))) %>%
  # create seasons
  mutate(season = case_when(date >= ymd("2015-11-01") & date <= ymd("2016-10-31") ~ "15/16",
                            date >= ymd("2016-11-01") & date <= ymd("2017-10-31") ~ "16/17",
                            date >= ymd("2017-11-01") & date <= ymd("2018-10-31") ~ "17/18",
                            date >= ymd("2018-11-01") & date <= ymd("2019-10-31") ~ "18/19",
                            date >= ymd("2019-11-01") & date <= ymd("2020-10-31") ~ "19/20",
                            TRUE ~ "NA")
  ) %>%
  
  # pre/early/mid/late-season
  mutate(training_phase = case_when(between(month, 2, 4) ~ 2,
                                    between(month, 5, 7) ~ 3,
                                    between(month, 8, 10) ~ 4,
                                    TRUE ~ 1)) %>%
  # convert workout duration from seconds to hours
  mutate(workout_time = workout_time/3600) %>%
  # remove seasons prior to 15/16
  filter(season != "NA")
```

Training season runs from 01/11 - 31/10 each year.

Training phases:
* Nov - Jan = 1 (pre-season)
* Feb - Apr = 2 (early-season)
* May - Jul 3 (mid-season)
* Aug - Oct = 4 (late season)

##### Data Distributions

```{r activities-per-season}
options(scipen = 999)

# View number of uploaded activities per season
activities_var %>%
  ggplot(aes(x = factor(season))) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Season",
       y = "Count") 

ggsave(path = "plots", filename = "activities_per_season.png")
  
```

```{r male-female}
activities %>%
  group_by(gender) %>%
  summarise(gender_count = n())

# As percentage
activities %>%
  group_by(gender) %>%
  summarise(count = n() ) %>%
  mutate(prop = count / sum(count) )

```

##### Filter Data

Create additional or mutate existing variables to assist in filtering and initial plotting



```{r filter-data}
activities_filtered <- activities_var %>%
  select(-X34) %>%
  # standard for research
  filter(age >= 18 & age <= 60) %>%
  # remove workout_time less than 10 seconds and more than 12 hours
  filter(workout_time >= 10/3600 & workout_time <= 12.0) %>%
  # remove speed > 65 kph (pro cyclist max sprint speed)
  filter(average_speed <= 65) %>%
  # indicate virtual power algorithm
  filter(`1s_critical_power` != 2000,
         `1s_critical_power` != 1000) %>%
  # remove absolute power values above 99th percentile for both genders
  pivot_longer(cols = `1s_critical_power`:`30m_critical_power`,
               names_to = "duration_s",
               values_to = "power_output") %>%
  group_by(gender,duration_s) %>%
  mutate(row = row_number()) %>%
  filter(power_output <= quantile(power_output, 0.99, na.rm = TRUE)) %>%
  pivot_wider(names_from = duration_s,
              values_from = power_output) %>%
  select(-row) %>%
  ungroup()

         
```

## Data Manipulation

##### Number of activites per individual

```{r number-of-rides-dists}
# how many rides do individuals do per /phase

## dummy column for cumulative count
activities_filtered$rec <- 1

activities_ride_count <- activities_filtered %>%
  arrange(date) %>%
  group_by(id) %>%
  mutate(ride_count = cumsum(rec)) %>%
  ungroup()


activities_total_rides <- activities_ride_count %>%
  # # rides per week
  # group_by(id, year, week) %>%
  # mutate(rides_per_week = sum(rec)) %>%
  # ungroup() %>%
  # rides per phase
  group_by(id, season, training_phase) %>%
  mutate(rides_per_phase = sum(rec)) %>%
  ungroup()

# how many rides do individuals upload in total?
activities_total_rides %>%
  ggplot(aes(x = n)) +
  geom_histogram(binwidth = 25) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  coord_cartesian(xlim = c(0, 2000)) +
  scale_x_continuous(breaks = seq(0,2000, by = 200)) +
  labs(title = "How many activites do individual users upload?",
       x = "Number of rides uploaded") 

# per phase histogram
activities_total_rides %>%
  group_by(id) %>%
  summarise(mean_rides_per_phase = mean(rides_per_phase)) %>%
  ungroup() %>%
  ggplot(aes(x = mean_rides_per_phase)) +
  geom_histogram(binwidth = 1) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  coord_cartesian(xlim = c(0, 250)) +
  scale_x_continuous(breaks = seq(0,250, by = 25)) +
  labs(title = "How many rides do people do per phase?",
       x = "Number of rides per phase")

# # per week histogram
# activities_total_rides %>%
#   group_by(id) %>%
#   summarise(mean_rides_per_week = mean(rides_per_week)) %>%
#   ungroup() %>%
#   ggplot(aes(x = mean_rides_per_week)) +
#   geom_histogram(binwidth = 1) +
#   theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
#   coord_cartesian(xlim = c(0, 20)) +
#   scale_x_continuous(breaks = seq(0,20, by = 1)) +
#   labs(title = "How many rides do people do per week?",
#        x = "Number of rides per week")



```

##### Phase best MMPs

```{r phase-best-mmp}
# Extract best MMP values for each yearly training phase
phase_mmps <- activities_filtered %>%
  group_by(id,season,training_phase) %>%
  summarise(across(c(`3m_critical_power`:`10m_critical_power`), max),.groups = "keep")

```


##### Critical power model

```{r critical-power-model}
# Convert to long format
phase_mmps_long <- phase_mmps %>%
  dplyr::rename(`180` = `3m_critical_power`,
                `300` = `5m_critical_power`,
                `480` = `8m_critical_power`,
                `600` = `10m_critical_power`) %>%
  pivot_longer(`180`:`600`, names_to = "duration_s", values_to = "power_output") %>%
  mutate(duration_s = as.numeric(duration_s),
    work_done = duration_s * power_output) %>%
  drop_na()

# Nest model data 
phase_mmps_nest <- phase_mmps_long %>%
  group_by(id,season,training_phase) %>%
  nest(data = c(duration_s,power_output,work_done))

# iterate linear cp model
cp_model <- phase_mmps_nest %>%
  mutate(model = map(data, ~lm(work_done ~ duration_s, data = .x)))

## iterate over the model with the function tidy (which puts lm output into tidy format) & unnest coeffs to extract cp & wprime into separate columns

activities_cp <- cp_model %>%
  rowwise() %>% 
  mutate(
    coefs = broom::tidy(model) %>% list()
  ) %>% 
  ungroup() %>% 
  select(id, season,training_phase ,coefs) %>% 
  unnest(cols = coefs) %>% 
  mutate(
    term = ifelse(term == "(Intercept)", "w_prime", "critical_power")
  ) %>% 
  select(id:std.error) %>% 
  pivot_longer(cols = c("estimate", "std.error")) %>% 
  mutate(
    term = ifelse(name == "estimate", term, paste(term, "error", sep = "_"))
  ) %>% 
  select(-name) %>% 
  pivot_wider(names_from = term) %>% 
  left_join(x = activities_filtered, 
            y = .,
            by = c("id", "season","training_phase")) %>% 
  ## Remove extreme wprime and CP values
  filter(w_prime > 1000,
         w_prime < 50000) %>%
  filter(critical_power > 100,
         critical_power < 500)

```

```{r power-duration-distribution}

ggplot(activities_cp,
       aes(x = critical_power)) +
  geom_histogram(binwidth = 5) +
  labs(title = "Critical Power distribution in males",
       x = "Critical Power (W)")

ggsave(path = "plots", filename = "cp_dist.png")
  

ggplot(activities_cp,
       aes(x = w_prime)) +
  geom_histogram(binwidth = 1000) +
  labs(title = "W' distribution in males",
       x = "W' (J)")

ggsave(path = "plots", filename = "wprime_dist.png")

```

##### Remove unwanted data frames

```{r}

```

##### Training phase metrics

```{r cp-quartiles + hw-pwr-ratio }

rides <- activities_cp %>%
  select(-contains(c("wpk", "_critical")), - data, -file,) %>%
  mutate(season = as.factor(season)) %>%
  mutate(training_phase = as.factor(training_phase)) %>%
  #mutate(week = as.factor(week)) %>%
  mutate(cp_quantile = ntile(critical_power,4)) %>%
  mutate(pwr_hr_ratio = average_power / average_hr) %>%
  mutate(relative_intensity = average_power / critical_power) %>%
  mutate(np_relative_intensity = coggan_np / critical_power) %>%
  mutate(np_hr_ratio = coggan_np / average_hr)
```


```{r calculate rides per week/phase}

# rides_weekly <- rides %>%
#   group_by(id, year, week) %>%
#   summarise(rides_per_week = n(), .groups = "keep")

rides_per_phase <- rides %>%
  group_by(id,season,training_phase) %>%
  summarise(rides_per_phase = sum(rec), .groups = "keep")

```


```{r cp-quartile-vis}
library(gghalves)
library(ggdist)

ggplot(rides, aes(x = as.factor(cp_quantile), y = critical_power)) +
  stat_halfeye(
    adjust = 0.5,
    width = 0.6,
    .width = 0,
    justification = -0.3) +
  geom_point(
    size = 1.3,
    alpha = 0.15,
    position = position_jitter(
      seed = 1, width = 0.1)
  ) +
  geom_boxplot(
    width = 0.25,
    outlier.shape = NA
  ) +
  labs(x = "Quartile",
       y = "Critical Power (W)")

ggsave(path = "plots", filename = "cp_quartile_boxplot.png")


```


```{r volume-by-week}
# # Weekly training metrics
# week_total_volume <- rides %>%
#   group_by(id,year,training_phase,week) %>%
#   summarise(across(c("workout_time_min","total_distance","elevation_gain","total_work", "coggan_tss"), ~sum(.x, na.rm = TRUE),.names = "week_{.col}"),.groups = "keep") %>%
#   dplyr::rename(week_total_workout_time = week_workout_time_min,
#                 week_total_elevation_gain = week_elevation_gain,
#                 week_total_tss = week_coggan_tss) %>%
#   ungroup()
# 
# 
# # Join week_total_volume with rides_weekly
# 
# week_total_volume_2 <- week_total_volume %>%
#   left_join(rides_weekly,by = c("id", "year","week"))
  
```

```{r intensity-by-week}
# week_avg_int <- rides %>%
#   # calculate relative session intensity
#   mutate(relative_intensity = average_power / critical_power) %>%
#   # calculate weekly averages
#   group_by(id,year,training_phase,week) %>%
#   summarise(across(c(average_speed:coggan_if,relative_intensity,pwr_hr_ratio), ~ mean(.x, na.rm = TRUE),.names = "week_{.col}"),.groups = "keep") %>%
#   ungroup()
```



```{r all-week-metrics}
# weekly_metrics <- full_join(week_total_volume_2, week_avg_int, by = c("id", "year", "training_phase", "week"))
```

```{r phase-metrics}

phase_totals <- rides %>%
  select(id, season, training_phase, workout_time, total_distance, elevation_gain, total_work) %>%
  group_by(id,season,training_phase) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE), .groups = "keep") %>%
  ungroup()

phase_totals <- full_join(phase_totals, rides_per_phase, by = c("id", "season", "training_phase"))

phase_averages <- rides %>%
  select(id,season, training_phase, average_speed, average_power, average_hr, average_cad, coggan_np, coggam_variability_index, coggan_if, coggan_tss, pwr_hr_ratio, np_hr_ratio,relative_intensity, np_relative_intensity) %>%
  group_by(id,season,training_phase) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE), .groups = "keep") %>%
  ungroup()

# merge

phase_metrics <- left_join(phase_totals,phase_averages, by = c("id", "season", "training_phase"))

```



```{r cp-metrics}

# add cp metrics to rides dataframe
cp_metrics <- rides %>%
  select(id,age,season,training_phase, gender,critical_power,w_prime,critical_power_error,w_prime_error,cp_quantile) %>%
  unique()

``` 


```{r merg-phase-and-cp}
phase_complete <- full_join(phase_metrics, cp_metrics, by = c("id","season","training_phase"))
```


###### Training metric plots

**Metrics by training phase**

Phase averages are calculated per training phase (3 month period)

```{r metrics-vis-1}
ggplot(phase_complete,
       aes(x = workout_time)) +
  geom_histogram(binwidth = 15) +
  labs(title = "Average training duration during",
       x = "Average training duration (hours)") +
  coord_cartesian(xlim = c(0,250))

ggsave(path = "plots", filename = "training_mins_dist.png")


```

```{r metrics-vis-2}

ggplot(phase_complete,
       aes(x = relative_intensity)) +
  geom_histogram(binwidth = 0.05) +
  labs(title = "Relative training intensity",
       x = "Proportion of critical power") 

ggsave(path = "plots", filename = "rel_int_dist.png")

```

```{r volume-by-quartile}
ggplot(phase_complete, aes(x = as.factor(cp_quantile), y = workout_time)) +
  stat_halfeye(
    adjust = 0.5,
    width = 0.6,
    .width = 0,
    justification = -0.3) +
  geom_point(
    size = 1.3,
    alpha = 0.15,
    position = position_jitter(
      seed = 1, width = 0.1)
  ) +
  geom_boxplot(
    width = 0.25,
    outlier.shape = NA
  ) +
  labs(title = "Training volume",
       x = "Quartile",
       y = "Average minutes per week")


ggplot(phase_complete, aes(x = workout_time, fill = as.factor(cp_quantile))) +
  geom_density(alpha = 0.5) +
  scale_colour_ggthemr_d() +
  labs(fill = "CP Quantile")

ggsave(path = "plots", filename = "vol_by_quartile.png")
```

```{r intensity-by-quartile}

ggplot(ride_metrics, aes(x = as.factor(cp_quantile), y = week_relative_intensity)) +
  stat_halfeye(
    adjust = 0.5,
    width = 0.6,
    .width = 0,
    justification = -0.3) +
  geom_point(
    size = 1.3,
    alpha = 0.15,
    position = position_jitter(
      seed = 1, width = 0.1)
  ) +
  geom_boxplot(
    width = 0.25,
    outlier.shape = NA
  ) +
  labs(title = "Training intensity",
       x = "Quartile",
       y = "Proportion of critical power")

ggplot(phase_complete, aes(x = relative_intensity, fill = as.factor(cp_quantile))) +
  geom_density(alpha = 0.5) +
  scale_colour_ggthemr_d() +
  labs(fill = "CP Quantile")

ggsave(path = "plots", filename = "rel_int_by_quartile.png")
```


```{r total-tss-quartile}
ggplot(phase_complete, aes(x = as.factor(cp_quantile), y = coggan_tss)) +
  stat_halfeye(
    adjust = 0.5,
    width = 0.6,
    .width = 0,
    justification = -0.3) +
  geom_point(
    size = 1.3,
    alpha = 0.15,
    position = position_jitter(
      seed = 1, width = 0.1)
  ) +
  geom_boxplot(
    width = 0.25,
    outlier.shape = NA
  ) +
  labs(title = "Training Stress Score Per Phase",
       x = "Quartile",
       y = "Training Stress Score (AU)")

ggplot(phase_complete, aes(x = coggan_tss, fill = as.factor(cp_quantile))) +
  geom_density(alpha = 0.5) +
  scale_colour_ggthemr_d() +
  labs(fill = "CP Quantile")

ggsave(path = "plots", filename = "total_tss_by_quartile.png")
```

**Summary table of training metrics**

```{r}
library(gt)
library(gtsummary)
library(webshot)  # save gt table as png

phase_metrics_table <- phase_complete %>%
  select(gender, critical_power ,workout_time, total_work, rides_per_phase) %>%
  dplyr::rename("Critical Power (W)" = critical_power,
                "Phase volume (hours)" = workout_time,
                "Work done per phase (kJ)" = total_work,
                "Rides per phase" = rides_per_phase) %>%
  tbl_summary(
    by = gender,
    missing = "no",
    statistic = all_continuous() ~ "{mean} ({sd})",
    digits = (all_continuous() ~ 1)) %>%
  as_gt %>%
  opt_align_table_header(align = "left") %>%
  tab_header(title = "Table 1. Training characteristics")

gtsave(phase_metrics_table, filename = "phase_training_metrics.png",path = "plots")
                
```

##### Results section plots/tables

```{r}
phase_complete %>%
  select(gender, critical_power ,workout_time, rides_per_phase,relative_intensity, w_prime) %>%
  filter(gender == "M") %>%
  dplyr::rename("Critical Power (W)" = critical_power,
                "W'" = w_prime,
                "Phase volume (hours)" = workout_time,
                "Rides per phase" = rides_per_phase,
                "Relative intensity" = relative_intensity) %>%
  tbl_summary(missing = "no",
    statistic = all_continuous() ~ "{mean} ({sd})",
    digits = (all_continuous() ~ 1)) %>%
  as_gt %>%
  opt_align_table_header(align = "left") %>%
  tab_header(title = "Table 1. Training characteristics")
```

```{r}
cp_dist <- phase_complete %>%
  filter(gender == "M") %>%
  ggplot(aes(x = critical_power)) +
  geom_histogram(binwidth = 10) +
  geom_vline(xintercept = 210.9, linetype = "dashed", colour = "#045DCE") +
  geom_vline(xintercept = 280.9, linetype = "dashed", colour = "#045DCE") +
  geom_vline(xintercept = 246.1, colour = "black") +
  labs(x = "Critical Power (W)",
       y = "Count")

ggsave(path = "plots", filename = "cp_dist_males.png")
```

```{r}
summary(phase_complete)
```



## Write to rds file
```{r}
write_rds(x = phase_complete,
          file = "data/ride_metrics.rds")
```




