---
title: "explore_2"
author: "Peter Bonner"
date: "25/02/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, warning = FALSE,
                      message = FALSE,echo = TRUE, dpi = 180,
                      fig.width = 8, fig.height = 5)


```

```{r libraries}
library(tidyverse)
library(readr)
library(janitor)
library(reshape)
library(readxl)
library(reshape2)
library(stringr)
library(lubridate)
# library(usethis)
# library(gitcreds)
theme_set(theme_minimal())

```

```{r}

# gitcreds_set()
# 
# use_git(message = "Initial commit")
# 
# use_github()
```




```{r}
activities_mmp <- read_csv("data/activities_mmp.csv")
```

The first part of the script uses the MMP data to create a key that meets the following critera:(https://gist.github.com/mpuchowicz/3b04ee65e97a4d6c32ee577a6cd9ea31)

1. Athlete has at least 100 power files for the year.
2. The MMP for the year is at least 7200 seconds in duration.
3. The MMP does not have obvious outlier data based on known virtual power issues and CP model parameters

```{r}
## I have re-organized the following because it is bad practice to
## re-assign a variable to the same name, unless you are doing this inside a function

# remove known problems. These two indicate virtual power algorithm
activities_mmp_filtered <- activities_mmp %>%
  filter(`1` != 2000) %>%
  filter(`1` != 1000) %>% 
  # remove values outside 99.9% percentile
  filter(`1` <= quantile(.$`1`,0.999, na.rm = TRUE)) %>% 
  filter(`180` <= quantile(.$`180`,0.999, na.rm = TRUE)) %>%
  filter(`420` <= quantile(.$`420`,0.999, na.rm = TRUE)) %>%
  filter(`720` <= quantile(.$`720`,0.999, na.rm = TRUE)) %>%
  filter(`1200` <= quantile(.$`1200`,0.999, na.rm = TRUE)) %>% 
  # remove spikey data
  filter(`1` < `5`*1.3) %>%
  filter(`1` < `2`*1.2) %>%
  filter(`1` != `2`)
```

The default value for format argument of as.Date function is "%Y-%m-%d" which is matching with the format of date column of df.

Hence, df$date <- as.Date(df$date) works perfectly. (I was getting NAs by using format = "%Y-%m-%d")

```{r}
library(lubridate)

# add a month, year and quarter columns
activities_mmp_dates <- activities_mmp_filtered %>%
  mutate(
    date = as_date(date),
    month = month(date),
    year = year(date),
    quarter = quarter(date)
  )
```

```{r}
# aggregrate year and quarterly bests
athlete_years <- activities_mmp_dates %>%
  group_by(id, year) %>%
  summarise(across(everything(), max)) %>% 
  ungroup()

athlete_quarters <- activities_mmp_dates %>%
  group_by(id, year, quarter) %>%
  summarise(across(everything(), max)) %>% 
  ungroup()

# add count of number of yearly rides per id
athlete_years_count <- activities_mmp_dates %>%
  group_by(id, year) %>%
  count() %>% 
  ungroup() %>% 
  left_join(athlete_years, .)

# sum all athlete ride duration

athlete_years_sum <- activities_mmp_dates %>%
  group_by(id, year) %>%
  summarise(across(where(is.numeric), sum)) %>% 
  ungroup() %>% 
  select(id, year, year_distance = distance, year_duration = duration) %>% 
  left_join(athlete_years_count, .)
```

```{r}
# get years with at least 52 rides (1 per week) & more than 120 mins total duration 

full_athlete_years <- athlete_years_sum %>%
  dplyr::rename(year_rides = n) %>%
  filter(year_rides >= 52, year_duration >= 7200)

# number of records (6277)
dim(full_athlete_years)

```

## Create CP Model

```{r}
library(tidymodels)

# Create long format data frame 

yearly_cp <- full_athlete_years %>%
  select(id, year, `180`, `420`,`720`) %>%
  pivot_longer(`180`:`720`, names_to = "trial_duration", values_to = "power_output") %>%
  mutate(trial_duration = as.numeric(trial_duration),
         work_done = trial_duration * power_output)


# Nest critical power trial data 
yearly_cp_nest <- yearly_cp %>%
  nest(data = -c(id, year))

# iterate over using linear CP model
yearly_cp_nest <- yearly_cp_nest %>%
  mutate(model = map(data, ~ lm(work_done ~ trial_duration, data = .x)))

#iterate over the model with the function tidy (which puts lm output into tidy format) & unnest coeffs to extract cp & wprime into separate columns

full_athlete_years_processed <- yearly_cp_nest %>%
  rowwise() %>% 
  mutate(
    coefs = broom::tidy(model) %>% list()
  ) %>% 
  ungroup() %>% 
  select(id, year, coefs) %>% 
  unnest(cols = coefs) %>% 
  mutate(
    term = ifelse(term == "(Intercept)", "w_prime", "critical_power")
  ) %>% 
  select(id:std.error) %>% 
  pivot_longer(cols = c("estimate", "std.error")) %>% 
  mutate(
    term = ifelse(name == "estimate", term, paste(term, "error", sep = "_"))
  ) %>% 
  select(-name) %>% 
  pivot_wider(names_from = term) %>% 
  left_join(full_athlete_years, .) %>% 
  ## Remove dodgy wprime and cp values
  filter(w_prime > 1000,
         w_prime < 50000) %>%
  filter(critical_power > 100,
         critical_power < 500)
```

## Data Visualisation

```{r}
## critical power distribution
full_athlete_years_processed %>%
  ggplot(aes(x = critical_power)) +
  geom_histogram(binwidth = 10) +
  geom_vline(aes(xintercept = mean(critical_power)),
                               colour = "blue",
                               linetype = "dashed")

```

```{r}
## wprime distribution
full_athlete_years_processed %>%
  ggplot(aes(x = w_prime)) +
  geom_histogram(binwidth = 1000)
```

```{r}
## plot power duration curves?

full_athlete_years_processed %>%
  select(id, year, c(`1`:`1800`)) %>%
  filter(year==2017) %>%
  pivot_longer(cols = `1`:`1800`,
               names_to = "duration",
               values_to = "power_output") %>%
  mutate(duration = as.numeric(duration)) %>%
  drop_na() %>%
  ggplot(aes(x = duration, y = power_output, group = id)) +
  geom_line(stat = "smooth",
            size = 0.5,
            alpha = 0.1,
            colour = "#f37b59") +
  ylim(0,2000) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

```{r}
# What summary statistics
library(gtsummary)

full_athlete_years_processed %>%
  group_by(year) %>%
  count()

## plot cp vs wprime

full_athlete_years_processed %>%
  ggplot(aes(x = critical_power, y = w_prime)) +
  geom_jitter(alpha = 0.3) +
  geom_smooth(method = "lm")

## long duration vs cp

full_athlete_years_processed %>%
  ggplot(aes(x = duration, y = critical_power)) +
  geom_jitter(alpha = 0.3)

```

