---
title: "explore_2"
author: "Peter Bonner"
date: "25/02/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(tidyverse)
library(readr)
library(janitor)
library(reshape)
library(readxl)
library(reshape2)
library(stringr)
install.packages("usethis")
library(usethis)

```

```{r}

use_github()
```




```{r}
activities_mmp <- read_csv("~/R/gc_open_data/data/activities_mmp.csv")
```

The first part of the script uses the MMP data to create a key that meets the following critera:(https://gist.github.com/mpuchowicz/3b04ee65e97a4d6c32ee577a6cd9ea31)

1. Athlete has at least 100 power files for the year.
2. The MMP for the year is at least 7200 seconds in duration.
3. The MMP does not have obvious outlier data based on known virtual power issues and CP model parameters

```{r}
# remove known problems. These two indicate virtual power algorithm

activities_mmp <- activities_mmp %>%
  filter(`1` != 2000) %>%
  filter(`1` != 1000)

# remove values outside 99.9% percentile
activities_mmp <- activities_mmp %>%
  filter(`1` <= quantile(activities_mmp$`1`,0.999, na.rm = TRUE)) %>%
  filter(`180` <= quantile(activities_mmp$`180`,0.999, na.rm = TRUE)) %>%
  filter(`420` <= quantile(activities_mmp$`420`,0.999, na.rm = TRUE)) %>%
  filter(`720` <= quantile(activities_mmp$`720`,0.999, na.rm = TRUE)) %>%
  filter(`1200` <= quantile(activities_mmp$`1200`,0.999, na.rm = TRUE)) 

# remove spikey data

activities_mmp <- activities_mmp %>%
  filter(`1` < `5`*1.3) %>%
  filter(`1` < `2`*1.2) %>%
  filter(`1` != `2`)

```

The default value for format argument of as.Date function is "%Y-%m-%d" which is matching with the format of date column of df.

Hence, df$date <- as.Date(df$date) works perfectly. (I was getting NAs by using format = "%Y-%m-%d")

```{r}
# add a month and year column
activities_mmp <- activities_mmp %>%
  mutate(date = as.Date(date)) %>%
  mutate(month = format(date, "%m")) %>%
  mutate(month = as.numeric(month)) %>%
  mutate(year = format(date, "%Y")) %>%
  mutate(year = as.numeric(year)) %>%
  mutate(quarter = quarter(date))

```

```{r}
install.packages("tidyselect")
library(tidyselect)

# aggregrate year and quarterly bests
athlete_years <- activities_mmp %>%
  group_by(id, year) %>%
  summarise(across(everything(), max))

athlete_quarters <- activities_mmp %>%
  group_by(id, year, quarter) %>%
  summarise(across(everything(), max))

# add count of number of yearly rides per id
athlete_years_count <- activities_mmp %>%
  group_by(id, year) %>%
  count()

# add count to athlete_years
athlete_years <- left_join(athlete_years, athlete_years_count, by = c("id", "year"))

# sum all athlete ride duration

athlete_years_sum <- activities_mmp %>%
  group_by(id, year) %>%
  summarise(across(where(is.numeric), sum))

athlete_years_sum <- athlete_years_sum %>%
  select(id, year, distance, duration) %>%
  dplyr::rename(year_distance = distance) %>%
  dplyr::rename(year_duration = duration)

athlete_years <-left_join(x = athlete_years, y = athlete_years_sum, by = c("id", "year"))


```

```{r}
# get years with at least 52 rides (1 per week) & more than 120 mins total duration 

full_athlete_years <- athlete_years %>%
  dplyr::rename(year_rides = n) %>%
  filter(year_rides >= 52) %>%
  filter(year_duration >=7200)

# number of records (6277)
dim(full_athlete_years)

```
```{r}
# use work-time model (provides 'middle' values) for CP & W'

## use 3 (180), 7 (420), 12 (720) MMP values to estimate CP and W' for each user ID per year - stuck here


```

